---
title: "Session 5: Solutions (Cirrhosis)"
format: html
editor: source
---

# Spider Angiomas and Platelet Count in Cirrhosis Patients

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

## Load Libraries and Import Data

We're bringing in cirrhosis data from the 1980's that was used for predictive modeling.  We'll use the `tidyverse` package to analyze our data.

Additional information about our data can be found at <https://archive.ics.uci.edu/dataset/878/cirrhosis+patient+survival+prediction+dataset-1>.

```{r libraries-ingest, warning = FALSE, message = FALSE}
library(tidyverse)
cirrhosis_study <- read_csv("../data/cirrhosis.csv") 
```

## Data cleaning

We're going to do some data cleaning, including converting Age to reflect years instead of days.

```{r clean-data}
cleaned_cirrhosis_study <- cirrhosis_study |>
  mutate(Status = factor(Status),
         Drug = factor(Drug),
         Age = Age/365.25,
         Sex = factor(Sex),
         Ascites = factor(Ascites),
         Hepatomegaly = factor(Hepatomegaly),
         Spiders = factor(Spiders),
         Edema = factor(Edema),
         Stage = factor(Stage))
```

## Show a Snapshot of the Data

Here's a peek at some of the statistical characteristics of the data:

```{r show-snapshot, echo=FALSE}
summary(cleaned_cirrhosis_study)
```

## Spider Angiomas 

We're going to concentrate on one discrete characteristic, the presence spider angiomas, recorded as either "Y" or "N" in the `Spiders` variable, and one continuous measurement, the number of thousands of platelets per cubic ml, stored in the `Platelets` variable.  We'll start with spiders angiomas.

Let's take a look at `Spiders` by looking at its prevalence per each cirrhosis disease stage.  We'll first make a slimmed down version of our data and remove rows for which we either don't know the stage of the disease or have no data about whether there are spiders.  Then we'll assess overall prevalence.

```{r spider-angioma-data, echo=FALSE}
spiders_complete_data <- cleaned_cirrhosis_study |>
  select(ID, Stage, Spiders) |>
  filter(!is.na(Stage),
         !is.na(Spiders))

spiders_prevalence_data <- spiders_complete_data |>
  group_by(Stage) |>
  summarise(spiders_pct = (sum(Spiders == "Y") / n()) * 100)

spiders_prevalence_data
```

OK, so it looks like the proportion of patients with spider angiomas increases with disease stage. Let's look at this visually!

```{r spiders-by-stage, echo=FALSE}
ggplot(spiders_complete_data) +
  geom_bar(aes(x=Stage, fill = Spiders),
           position = "dodge") 
```

A more publication-quality visualization:

```{r improved-spiders-by-stage, echo = FALSE}
ggplot(spiders_complete_data) +
  geom_bar(aes(x=Stage, fill = Spiders),
           position = "dodge") +
  labs(title = "Spider Angioma Prevalence Increases With Cirrhosis Severity",
       subtitle = "Data With Unknown Spider Angioma Status or Cirrhosis Stage Excluded",
       caption = "Data credit: UCI Machine Learning Repository",
       x = "Cirrhosis Disease Stage",
       y = "Participant Count") +
  scale_fill_manual(name = "Spider Angioma Status",
                    labels = c("N" = "Absent",
                               "Y" = "Present"),
                    values = c("N" = "goldenrod4",
                               "Y" = "darkorchid4"))
```

Finally, we can add labels right on the graph.  If our underlying data changes, our labels will, too!

```{r labeled-spiders-by-stage, echo = FALSE}
ggplot(spiders_complete_data) +
  geom_bar(aes(x=Stage, fill = Spiders),
           position = "dodge") +
  geom_label(data = spiders_prevalence_data,
             mapping = aes(x = Stage,
                           y = 70,
                           label = paste0("Stage ",
                                          Stage,
                                          ": ",
                                          round(spiders_pct), 
                                         "%\nPrevalence")),
             alpha = 0.8) +
  labs(title = "Spider Angioma Prevalence Increases With Cirrhosis Severity",
       subtitle = "Data With Unknown Spider Angioma Status or Cirrhosis Stage Excluded",
       caption = "Data credit: UCI Machine Learning Repository",
       x = "Cirrhosis Disease Stage",
       y = "Participant Count") +
  scale_fill_manual(name = "Spider Angioma Status",
                    labels = c("N" = "Absent",
                               "Y" = "Present"),
                    values = c("N" = "goldenrod4",
                               "Y" = "darkorchid4"))
```

## Platelets

OK, now let's look at `Platelet` data.  We'd like to see if there appears to be a relationship between platelet count and duration in study.  Let's also keep in mind that people left the study for three reasons: transplant, death, and end of the study.  These three endings or "status" designations might have different patterns, so let's look at them separately.

We're going to start by creating a cleaner dataset for just this question, with somewhat better labels for our `Status` variable.

```{r platelet-data, echo=FALSE}
platelet_data <- cleaned_cirrhosis_study |>
  select(ID,
         Platelets,
         N_Days,
         Status) |>
  mutate(Status = fct_recode(Status, 
                            "Censored" = "C", 
                            "Transplant" = "CL",
                            "Deceased" = "D"))

```

Let's first plot just the data points.

```{r platelet-plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(platelet_data) +
  geom_point(aes(x = Platelets, y = N_Days)) +
  facet_grid(vars(Status))
```

Now, let's plot both the data points themselves as well as the linear model that can be derived from the points, and add a bit of customization:

```{r, platelet-plot-with-model, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(platelet_data) +
  geom_point(aes(x = Platelets, y = N_Days),
             alpha = 0.7) +
  geom_smooth(aes(x = Platelets, y = N_Days),
              method = "lm",
              color = "darkgreen") +
  facet_grid(vars(Status)) +
  labs(title = "Increased Platelet Count Associated With Longer Study Duration",
       x = "Platelet Count per cubic ml/1000",
       y = "Study Duration in Days") +
  theme_bw() 
```

